---
title: "R Week 2 Assignment"
author: "Robert Godbey"
date: "July 8, 2015"
output: html_document
---
In probability, "n choose r" means n! / (( n - r)! * r!). For example, "5 choose 3" evaluates to 10. R has built in functions for choose() and factorial(). My task was to write my own variation of the choose() function, that does not use either of these functions.

My first step was to figure out how to multiply a series of numbers together from 1 to n, in order to get n factorial. I knew I could use a loop, but preferred not to use one. I found the prod() function and had my answer. To find 5 factorial I could use prod(1:5) or n! is prod(1:n).

This code snippet give us 2!, 3!, 4! and 5! (or 1 X 2, 1 X 2 X 3, 1 X 2 X 3 X 4, etc.)
```{r}
a <- 2
b <- 3
c <- 4
d <- 5

prod(1:a)
prod(1:b)
prod(1:c)
prod(1:d)

```

This allowed me to write n choose r in the following way in R. The following snippet gives us 5 choose 3 (which should equal 10 from above).

```{r}
n = 5
r = 3

nfac <- prod(1:n)           # This is n!
rfac <- prod(1:r)           # This is r!
nrfac <- prod(1:(n - r))    # This (n - r)!

mychoose <- nfac / (nrfac * rfac)
sprintf("%s choose %s = %s", n, r, mychoose)
```

Now all I needed to do was turn the above code into a function for any n and r and handle unexpected values for n and r. We can see from the equation above that both n and r need to be postive integers greater than zero and n needs to be greater than r. I discovered from testing I needed to improve my function to handle n = r, which should give the answer 1.

Here is the final function code for nchooser, my function for n choose r.

```{r}
# Function for "n choose r" means n! / (( n - r)! * r!) without choose() or factorial().

nchooser <- function(n, r) {
        # Let's start with some error handling.
        if (!is.numeric(n) || !is.numeric(r) || (n <= 0) || (r <= 0)){
                answer <- "This function requires two > 0 numbers to work."
        } else if (n < r) {
                answer <- "This function requires the first number > second."
        } else {
                # Multiply the numbers from 1 to x together to get n!, r!, and (n - r)!
                nfac <- prod(1:n)
                rfac <- prod(1:r)
                if (n == r) {
                        # This covers (n - r)! = 0! = 1
                        nrfac <- 1
                } else {
                        nrfac <- prod(1:(n - r))
                }
                mychoose <- nfac / (nrfac * rfac)
                answer <- sprintf("%s choose %s = %s", n, r, mychoose)
        }
        return(answer)
}
```


The test results for nchooser for these data follows. We can see the expected responses here.

- nchooser(5, 3) = 10
- nchooser(6, 3) = 20
- nchooser(6, 2) = 15
- nchooser(5, 5) = 1
- nchooser(3, 5) = error
- nchooser(5, 0) = error
- nchooser(5, "a") = error

```{r}
# Call the rchooser function with these test data

nchooser(5, 3)
nchooser(6, 3)
nchooser(6, 2)
nchooser(5, 5)
nchooser(3, 5)
nchooser(5, 0)
nchooser(5, "a")
```

These results convince me that my function is able to calculate n choose r.

